\documentclass{article}
\usepackage[utf8]{inputenc}
\input{header.tex}

\title{Diversity Statement}
\author{Caleb Stanford}
\date{March 2021}

\begin{document}

\maketitle{}

A strong commitment to diversity, equity, and inclusion is critical for the future of society and of our educational institutions.
In my years at Penn so far, I have had the opportunity to improve the diversity and climate at Penn through my leadership as co-chair of the Computer Science Doctoral Association (CISDA) since Fall 2018.
I have additionally served as a volunteer on the UPenn Applicant-Support Program (for under-served or under-represented communities in the admissions process), in Fall 2020.
Finally, I have been involved at the school (SEAS) level
through helping with the formation of the SEAS Dean's Advisory Board
and in meetings to establish the ``Our Values'' statement that
was highlighted on the SEAS website last year.
I hope that I will have more opportunities like these, and I will continue to listen to other students and researchers in other areas of academia, like education and gender and minority studies, to maintain progressive and informed thinking on these important issues.

The CIS Doctoral Association is a student-run advocacy organization for graduate students and for facilitating social engagement and community support in the CIS department.
One of its primary goals is to
provide a medium for student issues to be brought to the department,
through holding townhall meetings with the student body at least twice a semester and attending CIS faculty meetings.
We also regularly send out student surveys to gauge what issues students are facing on a larger scale, e.g. mental health issues students are facing during the COVID-19 pandemic.
As co-chair,
I have especially valued these opportunities to listen to the concerns of CIS students in the department and to represent these views as a representative to the faculty.
In addition to handling student issues, CISDA focuses on running student social events, such as weekly social hours and tea \& cookies.
In organizing such events I have worked with the women's doctoral association representatives to discuss ways in which we can make women students feel more included, and I have tried to facilitate a warm and welcoming community.
Through my time and efforts in CISDA, if I have only had a positive impact on the climate in the department, then that time and effort will have been worthwhile.

The UPenn Applicant Support Program was a new pioneering effort at UPenn in Fall 2020 to provide application support for minority and underrepresented students applying to PhD programs.
I was one of a dozen or two PhD student volunteers, and was paired with an applicant to the UPenn CIS PhD program with less access to useful resources and mentorship to help with their application.
I met with them several times throughout the semester and gave multiple rounds of feedback on their application materials.
I was delighted to hear that the student in question was admitted to Penn!
I found this to be a useful experience in improving my own ability
to function as a mentor, as well as to
revisit the various considerations in the application process.

At the SEAS level, I was involved in the early stages of organization of the SEAS Dean's Doctoral Advisory Board, which was set up to address student issues across the department, including issues of diversity and climate, especially advisor-student issues.
The Dean of SEAS additionally held a series of events to determine, as Penn Engineering, what our core values are and to release this in a public ``Our Values'' statement. I was fortunate to be involved in these meetings and to discuss and form the statement democratically with the other attendees.
One thing I learned from these meetings was that a principled and respectful approach is absolutely essential in order to listen to all parties involved, get faculty on board, and ultimately instigate change.
The selected values were: \emph{We are a community of scholarship, creativity and curiosity; We are a community of integrity and character; We are an inclusive community of diverse people and perspectives.}

I believe that student leadership organizations play a key role in the diversity and inclusion efforts at Penn.
Through all of these experiences and as a student leader, I have been fortunate to grow as a person and to bring student issues to the table.
I believe that it is imperative that we exercise intellectual humility in listening to and addressing the issues of disadvantaged communities,
just as we do in scientific research.

\end{document}
