Thank you to my wife, Ariana, for joining forces with me to complete our PhDs together over the last six years.
I've had your boundless support during the most exciting days, the shared relaxing evenings, and the hardest nights.
I'll miss exploring Philadelphia with you; I'm looking forward to the next adventure.

I am grateful to my advisor, Rajeev, for his exceptional patience, and for teaching me most of what I know about research.
Following our first paper submission, when I came to Rajeev with a crazy idea for a better way to define automata for streaming, he believed in me and let me take the project in that direction.
When I fixated on technical details, Rajeev never lost sight of the bigger picture, and that has been invaluable to my work.

I've been fortunate to have many other mentors at Penn. Thanks to Zack Ives, for his expertise in pragmatism; Benjamin Pierce, for teaching me how to think about programming languages and writing; Vincent Liu, for teaching me how to think about systems; and Val Tannen, who probably best understands what my research is trying to do (and always made the process of doing it thoroughly enjoyable).
Thank you to Stephanie Weirich and Steve Zdancewic for their advice on faculty interviews, for the board game night invites, and for taking the time to attend TGIF with students.
I also want to thank Kostas Mamouras, Scott Weinstein, and Sampath Kannan.
Thanks to my advisor at Amazon, Pauline Bolignano, for designing an internship which gave me close insight into what really matters to practitioners while still generating interesting research ideas; and to my advisors at Microsoft, Margus Veanes and Nikolaj Bjørner,
for being excited to talk about derivatives with me in spite of the pandemic, and for rapidly onboarding me on the Z3 project.

My life as a PhD student at Penn was made possible by the PLClub community.
Thanks to Leo Lampropoulos, Antal Spector-Zabusky, Jennifer Paykin, Robert Rand, and Christine Rizkallah for making me feel included when I was most in doubt;
to Konstantinos Kallas and Filip Nikšić, who made research noticeably more enjoyable after they joined, through many whiteboard conversations, dinners, and drinks;
and to several other friends, but especially Suguman Bansal, Kishor Jothimurugan, Yao Li, Solomon Maina, Omar Navarro Leija, Li-yao Xia, and Anton Xue.
More broadly, I have appreciated interacting with numerous members of the PLClub over the years:
Aalok, Antoine, Arthur, Calvin, Clara,
Elizabeth, Gautam, Halley, Haoxian, Hengchu,
Jiani, Kenny, Lawrence, Lei, Lucas,
Mukund, Nicolas, Nimit, Pardis, Pritam,
Rado, Richard, Steve, Sulekha, Xujie,
Yannick, Yishuai, and Ziyang.
% Arthur Azevedo de Amorim,
% Calvin Beck,
% Haoxian Chen,
% Pritam Choudhury,
% Elizabeth Dinella,
% Lawrence Dunn,
% Kenny Foner,
% Steve Hsu,
% Jiani Huang,
% Rado Ivanov,
% Nicolas Koh,
% Sulekha Kulkarni,
% Yishuai Li,
% Ziyang Li,
% Gautam Mohan,
% Pardis Pashakhanloo,
% Mukund Raghothaman,
% Clara Schneidewind,
% Lei Shi,
% Xujie Si,
% Lucas Silver,
% Nimit Singhania,
% Aalok Thakkar,
% Antoine Voizard,
% Halley Young,
% Yannick Zakowski,
% Hengchu Zhang,
% and
% Richard Zhang.
And best of luck to the next generation of PLClub stars
(though I know they don't need it):
these certainly include at least
Harry Goldstein, Phillip Hilliard, Stephen Mell, Irene Yoon,
Lef Ioannidis, Aaditya Naik,
Joe Cutler, Yiyun Liu, Jessica Shi, Joey Velez-Ginorio, and Chris Watson.

Besides PLClub, I owe a huge thanks to all of my fellow CISDA volunteers -- especially Luke Valenta, Daphne Ippolito, Oshin Agarwal, Paul He, and Alyssa Hwang -- for their efforts to make the department a better place.
Not to mention Cheryl Hickey, for single-handedly holding the department together.
My collaborators Nofel Yaseen, Liangcheng Yu, and Ryan Beckett have been very patient with my ignorance about networks.
I've enjoyed friendly conversations with Edo Roth, Zach Schutzman, Kelly Shiptoski, and Jiali Xing.
And thanks to Yu Chen for generously meeting with me to talk about online graph algorithms.
I'm sure I've forgotten someone, and I apologize!
Outside of Penn, I am grateful for the many conferences I have been able to attend, and the wonderful attendees I have met over the years.

Although my PhD has made too many of our interactions virtual, I can't overestimate the importance of my family and close friends.
Thank you to Kei Nishimura-Gasparian, Russell Shelp, and Frank Chung; my siblings, siblings-in-law, nieces, and nephews (MKZJTPT, JSSLCM, H, THL, BI, and E); my grandparents and extended family; Pinto and Calcifer;
and lastly, Alex Spentzos, for never running out of fun things to argue about, and for planning our wedding. In addition to Alex, my deep appreciation goes to Ariana's family, who welcomed me with open arms: Robin, Kyriakos, Sharon, Eric, Jasmin, Nathaniel, Michelle, Daniel, Aaron, Brian, and Cinder.

Some have sparked my love of thinking from the beginning.
Thank you, Mom and Dad (Kathleen and Joe);
Adella Croft, Michael Young, and Hiram Golze;
Canada/USA Mathcamp;
and Michael Zhao,
whose passion, intellect, and laughter I will always remember.
Thank you to my undergraduate mentors: Marion Scheepers, for a summer that made me certain that I wanted to do research as a career;
Tim Nelson, for getting me excited about logic for systems;
and Anna Lysyanskaya, who encouraged me and helped me apply to PhD programs in computer science seven years ago.
